Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-26 inference.tex
GKS-related edits.

The first-order encoding of a bind expression describes the semantics
of the set monad's bind operator in FOL. Let $s_1$ be a set, and let $f =
\lambda x.~s_2$ be a function that maps each variable in $s_1$ to a
new set. Then, $s = s_1 \bind f$ if and only if forall $x$, $y$, if
$x\in s_1$ and $y \in f(x)$, then  $y \in s$, and vice versa. The
translation first encodes $s_1$ and $s_2$ to obtain $(\phi_1,\G_1)$
and $(\phi_2,\G_2)$, respectively. Since $s_2$ is under a new binder
that binds $x$, the free variable sequence of $s_2$ is $\nubar,x$. In the
interest of hygiene, we substitute a fresh $a$ for $x$, making the
sequence $\nubar,a$. The set $s$ is encoded as a new uninterpreted
predicate $g$ indexed by $s$'s free variables ($\nubar$). Since the
set denoted by $g$ is the result of the bind $s_1 \bind \lambda
x.s_2$, a first-order constraint defining the bind operation (as
described above) is generated. The constraint relates the predicates
$\G_1$ and $\G_2$, representing $s_1$ and $s_2$ (resp.), to the
uninterpreted predicate $g$ that represents $s$. The constraint is
assigned a name ($\pi$) so that it can be enabled only for the choice
of free variables ($\nubar$) for which $g$ needs to be defined;
observe that the lambda expression returned by $\mssemof{\nubar}{s_1 \bind
\lambda x.s_2}$ asserts $\pi(\vbar)$ only in conjunction with
$g(\vbar,r)$, for some $r$.

The first-order encoding of the $\existsl(\stg,\phi,s)$ expression
eliminates the existential by Skolemizing it. Skolemizing is the
process of substituting an existentially bound $x$ in
$\phi_x\in\Prop^1$ with $f(\nubar)$, where $f$ is a fresh
uninterpreted function, and $\nubar$ are the free variables in
$\phi_x$ bound by enclosing universal quantifiers. Due to the
decidability restrictions (Sec.~\ref{sec:decidability}), the only
uninterpreted functions we admit in our logic are boolean (i.e.,
predicates/relations). Consequently, Skolemizing an existentially
bound $\stg$ with $f(\nubar)$ is not really an option in our case.
Instead, we define $f$ as an uninterpreted relation that behaves as a
partial function:
\begin{smathpar}
  \begin{array}{c}
    \forall \nubar.\forall a.\forall b.~f(\nubar,a) \wedge f(\nubar,b)
    \Rightarrow a = b
  \end{array}
\end{smathpar}
We then replace the existentially bound $\stg$ with a new universally
bound $a$ in $\phi$ and $s$, and define the set $[a/\stg]s$ under the
condition that $f(\nubar,a) \conj [a/\stg]\phi$ holds. Since the
existence of a $\stg$ that satisfies $\phi$ is axiomatized to
define $s$, if the aforementioned condition does not hold, i.e., if
there does exist such a $\stg$, we let the context derive $\bot$.

@@

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a fragment of FOL that satisfies the following syntactic properties:
\begin{itemize}
  \item All function symbols, modulo those that might appear in
    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.
\end{itemize}
The simple syntactic structure of the fragment already makes is
amenable for automatic reasoning via an off-the-shelf SMT solver, such
as Z3. The decidability of this fragment, however, is more subtle and
discussed below.

Consider a set expression $s$ with no free variables (i.e., $\nubar =
\emptyset$, like $\T(\stl,\stg)$ from the above example). Let
$(\phi,G) = \mssemof{\emptyset}{s}$. Note that $\phi$ is a conjunction
of (a).  $\phi_i$'s, where each $\phi_i$ results from encoding a
subexpression $s_i$ of $s$, and (b). a $\phi_s$, resulting from
encoding $s$ itself (i.e., its top-level expression). From Fig.~\ref{fig:logic},
it is clear that $\phi_s$ is either $\top$ (for the first four cases),
or it is a prenex-quantified formula, where quantification is either
$\forall^2$, or $\exists$, or $\forall\exists$. Generalizing this
observation, for a set expression $s$ with $|\nubar|$ free variables,
$\phi_s$, if quantified, is a prenex-quantified formula, where
quantification assumes one among the forms of $\forall^{|\nubar|+2}$,
or $\forall^{|\nubar|}\exists$, or $\forall^{|\nubar|+1}\exists$.  In
other words, the number of $\forall$ quantifiers preceding an
$\exists$ quantifier is utmost two more than the number of free
variables ($\nubar$) in $s$. For the convenience of this discussion,
let us call $\forall^{|\nubar|+1}\exists$ as the prenex signature of
$\phi_s$.

Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
is extended only in the encoding rules for $\bind$ and $\existsl$
expressions. Since $\existsl$ is a result of a particular choice of the
weakening semantics (Sec.~\ref{sec:inference}), let us ignore it for now to
focus on the core set expression language. This leaves $\bind$ as the
only case where $\nubar$ is extended.  Since an occurrence of $\bind$
adds one to $|\nubar|$, if $s$ is a subexpression of a top-level
{\sf bind} expression (like $\F_U(\stl,\stg)$ from the above example),
then the prenex signature of $\phi_s$ is $\forall^2\exists$.
Furthermore, if the subexpressions of $s$ are not {\sf bind} expressions,
then none of the $\phi_i$'s are quantified, and the prenex signature
of $\phi = \bigwedge_i\phi_i \wedge \phi_s$ remains
$\forall^2\exists$.  The fragment of FOL that admits formulas with
prenex signatures of form $\forall^2\exists^*$ is called the
Godel-Kalmar-Schutte ({\sf GKS}) fragment~\cite{...}, which is known to be
decidable. The language of encoding, however, is a combination of {\sf
GKS} with (a). $\Prop^0$, the theory from which quantifier-free
propositions ($\varphi$) that encode object language expressions are
drawn, and (b).  $\Prop^1$, the theory from which invariants ($I$) are
drawn. Thus the encoding is decidable if the combination of ${\sf GKS}
+ \Prop^0 + \Prop^1$ is decidable.

To formally state the above property, let $\SL[n,\Prop^0,\Prop^1]$
denote the core set expression language that admits $n$ nested {\sf bind}
expressions, and allows program (Boolean) expressions ($e$), and
invariants ($\I$) to be drawn from theories $\Prop^0$ and $\Prop^1$,
respectively. Then:
\begin{theorem}
  $\SL[0,\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
The $n=0$ case denotes the lack of nested {\sf bind}
expressions. Since nested \C{foreach} statements are the only program
constructs in $\txnimp$ that lead to nested bind expressions, this
instantiation denotes the encoding language for $\txnimp$ programs
without nested \C{foreach} loops. Observe that the \C{new\_order}
transaction (Fig.~\ref{fig:new_order_code}) satisfies this constraint.
Indeed, this restricted language is nonetheless expressive enough to
cover all of the benchmarks we have considered.

>>>>

The first-order encoding of a bind expression describes the semantics
of the set monad's bind operator in FOL. Let $s_1$ be a set, and let
$f$ be a function that maps each variable in $s_1$ to a new set. Then,
$s_2 = s_1 \bind f$ if and only if for all $y\in s_2$, there exists an
$x \in s_1$ such that $y = f(x)$, and forall $x\in s_1$, $f(x)\in
s_2$. The encoding essentially adds new constraints to this effect.
The translation first encodes $s_1$ and $s_2$ to obtain
$(\phi_1,\G_1)$ and $(\phi_2,\G_2)$, respectively. Since $s_2$ is
under a new binder that binds $x$, the free variable sequence of $s_2$
is $\nubar,x$. In the interest of hygiene, we substitute a fresh $a$
for $x$, making the sequence $\nubar,a$. The set $s$ is encoded as a
new uninterpreted predicate $g$ indexed by $s$'s free variables
($\nubar$). Since the set denoted by $g$ is the result of the bind
$s_1 \bind \lambda x.s_2$, first-order constraints defining the bind
operation (as described above) are generated. The constraints relate
the predicates $\G_1$ and $\G_2$, representing $s_1$ and $s_2$
(resp.), to the uninterpreted predicate $g$ that represents $s$. The
constraints are assigned names ($\pi_1$ and $\pi_2$) to give them an
easy handle.

The first-order encoding of the $\existsl(\stg,\phi,s)$ expression
essentially Skolemizes the existential. Skolemizing is the
process of substituting an existentially bound $x$ in
$\phi_x\in\Prop^1$ with $f(\nubar)$, where $f$ is a fresh
uninterpreted function (called the Skolem function), and $\nubar$ are
the free variables in $\phi_x$ bound by enclosing universal
quantifiers. Due to the
decidability restrictions (Sec.~\ref{sec:decidability}), the only
uninterpreted functions we admit in our logic are boolean (i.e.,
predicates/relations). Consequently, we cannot define the Skolem
function $f$ directly. Instead, we define it via an uninterpreted
relation, by explicitly asserting the function property:
\begin{smathpar}
  \begin{array}{c}
    (\forall \nubar.\forall a.\forall b.~f(\nubar,a) \wedge f(\nubar,b)
    \Rightarrow a = b)
    ~~\conj~~ (\forall \nubar.\exists a. f(\nubar,a))
  \end{array}
\end{smathpar}
We then replace the existentially bound $\stg$ with a new universally
bound $a$ in $\phi$ and $s$, such that $f(\nubar,a) $ holds, before
encoding the existentially bound $s$.

@@

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a fragment of FOL that satisfies the following syntactic properties:
\begin{itemize}
  \item All function symbols, modulo those that are drawn from
    $\Prop^0$ and $\Prop^1$, are uninterpreted and Boolean.
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.
\end{itemize}
The simple syntactic structure of the fragment already makes is
amenable for automatic reasoning via an off-the-shelf SMT solver, such
as Z3. The decidability of this fragment, however, is more subtle and
discussed below.

Consider a set expression $s$ with no free variables (i.e., $\nubar =
\emptyset$, like $\T(\stl,\stg)$ from the above example). Let
$(\phi,G) = \mssemof{\emptyset}{s}$. Note that $\phi$ is a conjunction
of (a).  $\phi_i$'s, where each $\phi_i$ results from encoding a
subexpression $s_i$ of $s$, and (b). a $\phi_s$, resulting from
encoding $s$ itself (i.e., its top-level expression). From Fig.~\ref{fig:logic},
it is clear that $\phi_s$ is either $\top$ (for the first four cases),
or it is a prenex-quantified formula, where quantification is either
$\forall^2$, or $\exists$, or $\forall\exists$. Generalizing this
observation, for a set expression $s$ with $|\nubar|$ free variables,
$\phi_s$, if quantified, is a prenex-quantified formula, where
quantification assumes one among the forms of $\forall^{|\nubar|+2}$,
or $\forall^{|\nubar|}\exists$, or $\forall^{|\nubar|+1}\exists$.  In
other words, the number of $\forall$ quantifiers preceding an
$\exists$ quantifier is utmost one more than the number of free
variables ($\nubar$) in $s$. For the convenience of this discussion,
let us call $\forall^{|\nubar|+1}\exists$ as the prenex signature of
$\phi_s$.

Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
is extended only in the encoding rule for $\bind$. Since an occurrence
of $\bind$ adds one to $|\nubar|$, if $s$ is a bind expression nested
inside a top-level bind expression (like $\F_U(\stl,\stg)$ from
the above example), then the prenex signature of $\phi_s$ is
$\forall^2\exists$.  Furthermore, if the subexpressions of $s$ are
neither {bind} nor $\existsl$ expressions, then none of the
$\phi_i$'s are quantified, and the prenex signature of $\phi =
\bigwedge_i\phi_i \wedge \phi_s$ remains $\forall^2\exists$. Similar
observation holds when $s$ is an $\existsl$ expression nested inside a
top-level {bind} expression.  Since $\existsl$ is generated as a
result of stabilizing a SQL command transformer, which is always a
non-nested bind expression, the subexpression ($s'$) of $\existsl$ is
a non-nested bind expression. $s'$ is however nested inside the
top-level bind expression, hence its prenex signature is
$\forall^2\exists$.  Since $\existsl$ doesn't extend $\nubar$, the
prenex signature of $s$ remains $\forall^2\exists$. When $s$ is an
expression other than $\bind$ or $\existsl$, then $\phi_s$ is not a
quantified formula, and its prenex signature is trivially subsumed by
$\forall^2\exists$. Thus, for the subset of $\SL$, where bind
expressions are restricted to one level of nesting, the FOL formulas
generated by the encoding have the prenex signature as
$\forall^2\exists$.

The fragment of FOL that admits formulas with prenex signatures of
form $\forall^2\exists^*$ is called the G\"odel-K\'almar-Sch\"utte ({\sf
GKS}) fragment~\cite{gks}, which is known to be decidable. The
language of encoding, however, is a combination of {\sf GKS} with (a).
$\Prop^0$, the theory from which quantifier-free propositions
($\varphi$) that encode object language expressions are drawn, and
(b). $\Prop^1$, the theory from which invariants ($I$) are drawn. Thus
the encoding of the subset of $\SL$ described above is decidable if the
combination of ${\sf GKS} + \Prop^0 + \Prop^1$ is decidable. We write
$\SL[\Prop^0,\Prop^1]$ to highlight the parameterization of $\SL$ on
$\Prop^0$ and $\Prop^1$.  The discussion in the previous paragraph
points to the existence of subsets in $\SL[\Prop^0,\Prop^1]$ that are
decidable:
\begin{theorem}
  There exists an $\SL'[\Prop^0,\Prop^1] \subset \SL[\Prop^0,\Prop^1]$
  such that $\SL'$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
  decidable.
\end{theorem}
An example of such an $\SL'$ is the subset described above: $\SL$ with
bind expressions confined to one level of nesting. We denote this
subset as $\SL^1[\Prop^0,\Prop^1]$, for which we assert decidability:
\begin{theorem}
  $\SL^1[\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
  decidable.
\end{theorem}
$\SL^1$ is a useful subset of $\SL$, for it corresponds to $\txnimp$
programs without nested \C{foreach} loops. Observe that the
\C{new\_order} transaction (Fig.~\ref{fig:new_order_code}) belongs to
this subset.  Indeed, $\SL^1$, while being a restricted version of
$\SL$, is nonetheless expressive enough to cover all the benchmarks we
considered in Sec.~\ref{sec:case-studies}.

A useful instantiation of $\SL^1$ is $\SL[{\sf BV},{\sf GKS}+{\sf
BV}]$, where ${\sf BV}$ is the theory of bit-vector arithmetic, which
is often used to encode the finite-bit integer arithmetic of real
programs. Finite-bit integer arithmetic has a finite axiomatization in
{\sf GKS}. For instance, 32-bit integers can be encoded as $2^{32}$
distinct constants of an uninterpreted sort $T$, while integer
operations like addition and multiplication can be encoded as
uninterpreted functions whose properties are enumerated for the entire
domain of $T$. Thus {\sf BV} is subsumed by {\sf GKS}. Since the
latter is decidable, the combination is decidable:
\begin{theorem}
  $\SL^1[{\sf GKS},{\sf GKS}+{\sf BV}]$ is decidable.
\end{theorem}
The $\SL^1[{\sf BV},{\sf GKS}+{\sf BV}]$ instantiation requires $I$ to
be drawn from {\sf GKS}+{\sf BV}, which is expressive enough to
describe common database integrity constraints, such as referential
integrity, non-negativeness of all integer values in a column etc.
The isolation specifications presented in \S\ref{sec:isolation} are
already simple first-order formulas that can be encoded in {\sf GKS}.
Furthermore, it is also reasonable to expect the guarantee ($G$) of a
transaction to be expressible in the same logic as its inferred $\F$,
since $\F$ (without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL^1[{\sf GKS},{\sf GKS}+{\sf BV}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable.
